I would like to express my deepest gratitude to my advisor, Yannis Smaragdakis, for being such a great mentor to me, throughout my PhD studies and before. I immensely appreciate his endless encouragement, patient, and motivation; his advice and guidance have been priceless---even during passionate scientific arguments.

I also thank Alex Delis, Panos Rondogiannis, Mema Roussopoulos, Manolis Koubarakis, Panagiotis Stamatopoulos, and Nikolaos Papaspyrou for their valuable comments and advice while serving as members of my dissertation committee.

My sincere thanks to Shan Shan Huang, Martin Bravenboer, and Molham Aref, who gave me the opportunity to join LogicBlox as an intern, during my PhD. Furthermore, I also extend my thanks to Ben Livshits who gave me the opportunity for an internship in Microsoft Research, and later guided me as a mentor throughout the process. Both internships proved memorable experiences and I am indebted to all those who make them a reality.

A special thanks to my labmates throughout the years: Aggelos Biboudis, George Balatsouras, George Fourtounis, George Kollias, Anastasios Antoniadis, Kostas Saidis, Petros Pathoulas, Neville Grech, Christos Vrachas, Sifis Lagouvardos, Kostas Ferles, Efthymios Hadjimichael, Dimitris Galipos, Stamatis Kolovos, Konstantinos Triantafyllou and Ilias Tsatiris. I am grateful for our interactions, discussions, and potential arguments over every possible aspect of programming languages that intrigued us to no end, and for all the fun we had.

Finally, I would like to thank Anna for always being positive and understanding and my parents, Stavros and Aristea, for being so supportive and reassuring over all these years.